---
title: "Import and Explore"
author: "Eoin Scanlon"
date: "2/5/2020"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)

```

## Import and sneak peak

Data can be downloaded from <https://www.kaggle.com/c/dont-overfit-ii/data>. Always good to have a look at the structure of the data first.

```{r}
train <- read.csv(file = 'train.csv')
#head(train)
ncol(train)
nrow(train)
```

The column 'target' is the outcome that we will wish to predict, and the rest are just arbitrary, meaningless data. The training dataset is much, much smaller than the test. Interesting. Lets see some summary statistics of the data.

```{r}
#summary(train)
```

This is very hard to read, so lets see how to interpret this a bit better.

```{r}
train.data <- train[,-c(1,2)] # Remove unnecessary columns
train.means <- colMeans(train.data) # Means of each column
train.sd <- apply(train.data, 2, sd) # SD of each column

ggplot(as.data.frame(train.means), aes(x = train.means)) + geom_histogram(color="black", fill="blue",binwidth = 0.01)
ggplot(as.data.frame(train.sd), aes(x = train.sd)) + geom_histogram(color="black", fill="green",binwidth = 0.01)

```


